---
title: "Tidyverse Introduction"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(plotly)
dat <- suppressMessages(readr::read_csv("data/dat.csv", na = "."))
dat2 <- suppressMessages(readr::read_csv("data/dat2.csv", na = "."))
dat3 <- suppressMessages(readr::read_csv("data/dat3.csv", na = "."))

knitr::opts_chunk$set(echo = FALSE)
```


## Tidy data with tidyr

A nice spreadsheet with colors and multiple tables on one sheet might be readable by humans, but is not very compatible with computers and data analysis.

Even a single table can be awkward if it is not **tidy**.

### What are tidy data?

1. Each **variable** is in its own column, and each column is a variable
2. Each **observation** is in its own row and each row is an observation
3. Each **value** is a cell and and each cell is a single value


Write the R code to examine 3 tables: table1, table2, and table3.

```{r tables, exercise=TRUE}

```

```{r tables-solution}
table1
table2
table3
```

```{r tidy-table}
question("Now, which of the tables is tidy?",
         answer("table1", correct = TRUE),
         answer("table2"),
         answer("table3"),
         incorrect = "Not quite. The type column in table2 has has two variables (rule 1), and observations split into two rows (rule 2). The rate column of table3 has more than one value per cell (rule 3).",
         correct = "Yes, table1 is tidy. The type column in table2 has has two variables (rule 1), and observations split into two rows (rule 2). The rate column of table3 has more than one value per cell (rule 3)."
)
```

Look at this table. This is the format for pharmacometric data sets used by the [Pmetrics](https://github.com/LAPKB/Pmetrics){target="_blank"} R Package. It is similar to the data structure for most pharmacometric modeling software tools.

`r dat2`

```{r tidy-pmtable}
question("Is it tidy?",
         answer("Yes"),
         answer("No", correct = TRUE),
         incorrect = "Not quite. It is almost tidy, but subject 1 for example has two observations at time 120 that are in two rows. One way to be tidy is to put the dose value and out value for time 120 in the same row.",
         correct = "That's right. It is almost tidy, but subject 1 for example has two observations at time 120 that are in two rows. One way to be tidy is to put the dose value and out value for time 120 in the same row."
)
```

<details>
  <summary>More on tidy tables</summary>
  Another way to make the table tidy is with an additional `EVID` column.
  
  `r dat3`
  
  ```{r tidy-pmtable1}
  question("Now is it tidy?",
          answer("Yes", correct = TRUE),
          answer("No"),
          incorrect = "It is in fact tidy, because the EVID column now means that each row is a distinct event and each event is in a distinct row.",
          correct = "Yes, it has become tidy, because the EVID column now means that each row is a distinct event and each event is in a distinct row."
          )
  
  ```
</details>

### Functions to help tidy data

Most of the functions to help make data tidy are in the [tidyr](https://tidyr.tidyverse.org){target="_blank"} package.

The most helpful functions I've used are

* pivot_longer
* fill

Let's look at examples of these.

#### Pivot Longer

```{r pivot-longer, exercise = TRUE}
table4a
table4a |> pivot_longer(cols = !country, names_to = "year", values_to = "cases")
```

Do you see how `table4a` became longer? It didn't become narrower, but that can happen. There is a companion function `pivot_wider` which makes tables wider of course, but it is used less often. There are some notes to make about the code above.

* We used a **pipe** `|>`. This operator feeds the left side into the right side, supplying the results of the left side as the initial function argument for the right side. Base R has adopted this pipe, inspired by the original pipe `%>%` in the `magrittr` package. Either pipe will work, but there are some subtle differences for advanced users. Pipes are an incredibly useful way to build a series of data transformations. The first argument for `pivot_longer` is actually `data`, but we don't have to specify that because we piped our data to `pivot_longer`. We'll see in a moment how to use a series of pipes.
* The next argument is `cols`, which tells `pivot_longer` which columns contain the data which will be made longer. We used `!` to select all columns *except* country, but check out all the ways you can choose columns for pivoting with [tidy-select](https://tidyr.tidyverse.org/reference/tidyr_tidy_select.html){target="_blank}. 
* The next argument is `names_to` which will be the name of the column created by extracting the information in the names of the columns specified by `cols` to pivot.
* Lastly, we used `values_to` to name the column that contains the data originally in the `cols` columns.

#### Fill

Look at a pharmacometric data table for the [Pmetrics](https://lapkb.github.io/Pmetrics/){target="_blank"} R Package.

```{r echo = TRUE}
dat |> select(ID, TIME, WT, AFRICA, AGE, GENDER, HEIGHT)
```

We use the pipe again plus `select`, a function from the `dplyr` package which we'll look more at later. This helps us focus on the columns of interest. 

Notice how the values for the covariates are only present in the first row at time 0 for each subject. Let's fill in the missing values.

```{r echo = TRUE}
dat |> select(ID, TIME, WT, AFRICA, AGE, GENDER, HEIGHT) |> fill(WT)
```

We used two pipes serially to accomplish this. We have not modified the original `dat` object. We simply piped it into two functions that first selected columns of interest, and then filled in missing values for WT.

Try writing piped code on your own that:

1. Creates a new object called `dat_fill`
2. Selects the ID, TIME, WT, AGE, GENDER and HEIGHT columns only
3. Fills in the values for WT, AGE, GENDER and HEIGHT.

```{r fill-exercise, exercise = TRUE}

```

```{r fill-exercise-solution}
dat_fill <- dat |> select(ID, TIME, WT, AGE, GENDER, HEIGHT) |> fill(WT, AGE, GENDER, HEIGHT)
dat_fill #prints the result
```

What happens if you use the <tidy-select> `everything()` function as the only argument to `fill()`? Pay attention to the behavior.

## Data manipulation with dplyr

Tidy data are easier to work with. There are many functions to help you slice and dice it. Most of these functions are within the `dplyr` package, part of the Tidyverse.

Let's look at these very commonly used and useful functions:

* **Summarize**
  * summarize()
  * count()
* **Group**
  * group_by() / ungroup()
  * across()
* **Order**
  * arrange()
* **Extract**
  * filter()
  * distinct()
  * slice()
  * select()
* **Create**
  * mutate()
  * rename()
  * bind_rows() / bind_cols()

Let's knock off a few of these at once. We can use `summarize` to generate one-row summaries of columns. Add `across` to vectorize the summary across multiple columns.

```{r echo = TRUE}
dat |> summarize(across(WT:HEIGHT, \(x) mean(x, na.rm = TRUE)))
```

Let's break that down. We pipe `dat` into the `summarize` function. Instead of just getting the mean of one column, we use the `across` function to specify the columns in `dat` we want to summarize (WT:HEIGHT), then use a "lambda" function on the fly to get the mean of each column: `\(x) mean(x,na.rm = TRUE)`. This is the most current way to specify a lambda function in Tidyverse packages. An older way, which still works, is `~ mean(.x, na.rm = TRUE)`. Note the `.x` instead of `x`. The construct with `\(x)` is preferable, because you could just as easily use `\(y)` or any other indexing variable. Note that we have to tell `mean()` to ignore missing values (`NA`) with the `na.rm = TRUE` argument to `mean()`.

You try summarizing the median of WT and AGE only.
```{r across, exercise = TRUE}

```

```{r across-solution}
dat |> summarize(across(c(WT,AGE), \(x) median(x, na.rm = TRUE)))
```

Let's combine another couple of functions. How many unique subjects are there?

``` {r echo = TRUE}
dat |> distinct(ID) |> count() #tidyverse
length(unique(dat$ID)) #base R
```

What if we want the mean drug concentration by subject, but only for the first 4 and last 4 subjects? That's where `filter` and `group_by` come in handy.

``` {r echo = TRUE}
dat |> filter(ID <= 4 | ID >=17) |> group_by(ID) |> summarize(mean_conc = mean(OUT, na.rm = TRUE)) 
```

Now, try finding the first dose for each subject. Hint, `group_by` and `slice_head` will be helpful here. You should only return the ID and DOSE columns by using `select` as you did earlier.

```{r slice, exercise = TRUE}

```

```{r slice-solution}
dat |> group_by(ID) |> slice_head(n = 1) |> select(ID, DOSE)
```

You can use `arrange()` to change the order. Add to your last statement to arrange first by DOSE, then by ID.

```{r arrange, exercise = TRUE}

```

```{r arrange-solution}
dat |> group_by(ID) |> slice_head(n = 1) |> select(ID, DOSE) |> arrange(DOSE, ID)
```

What if we want to combine variables in some fashion to make a new variable? Let's try making a new column called "bmi" for body mass index. The formula is $kg/m^2$. We'll use the `mutate()` function for that.

``` {r echo = TRUE}
dat |> mutate(bmi = WT/(HEIGHT/100)^2) #HEIGHT is in cm, so need to convert to meters
```

Click the black triange next to OUTEQ to scroll the table and see your new bmi variable. Since the table also has a lot of empty rows in it, we can generate a summary. While we looked before at creating a single summary statistic with `summarize`, you can also pipe to functions in other R packages, as long as the first argument is the data to be acted upon. Check this out.

``` {r echo = TRUE}
dat |> mutate(bmi = WT/(HEIGHT/100)^2) |> summary(.x)
```

We use the `.x` notation to send the output of the last pipe to the base R `summary` function. The nice thing about this function is that it gives us several statistics all at once, and automatically removes the `NA` values, telling us how many there were.

OK, let's try to put it together into a more complex statement. Try the following pipe steps on your own.

1. Select the ID, TIME, DOSE, and OUT columns from `dat`
2. Group by ID
3. Create new variable called norm_out for OUT normalized to the mean dose for each subject. Hint, don't forget to use `na.rm = TRUE` when you calculate the mean dose for each subject.
4. Filter out the records that are missing for norm_out. Hint, use `is.na()` in your filter, negated. Remember how to negate from the exercise with `pivot_longer` above?
5. Drop the DOSE column.
6. Summarize the mean of norm_out for each subject and call it mean_norm_out. 

```{r norm-out, exercise = TRUE}

```

```{r norm-out-solution}
dat |> select(ID, TIME, DOSE, OUT) |> group_by(ID) |> mutate(norm_out = OUT / mean(DOSE, na.rm = TRUE)) |> filter(!is.na(norm_out)) |> select(!DOSE) |> summarize (mean_norm_out = mean(norm_out))
```

What if you wanted to just get the mean dose normalized output over the whole population? You can insert `ungroup()` into your pipeline just before the summarize statement.

```{r norm-out-ungroup, exercise = TRUE}

```

```{r norm-out-ungroup-solution}
dat |> select(ID, TIME, DOSE, OUT) |> group_by(ID) |> mutate(norm_out = OUT / mean(DOSE, na.rm = TRUE)) |> filter(!is.na(norm_out)) |> select(!DOSE) |> ungroup() |> summarize (mean_norm_out = mean(norm_out))
```

If you did everything correctly, you should have found a value of 0.0125 for the mean of each subject's mean dose-normalized concentrations.

Say you're familiar with another modeling package that calls their "OUT" values as "DV" for dependent variable. It's easy enough with Tidyverse to rename that column.

```{r echo = TRUE}
dat |> rename(DV = OUT) |> slice_head(n = 10) #first 10 rows just for conciseness
```

The format for `rename` is "new = old". You can include as many pairs as you'd like. There's even an nice variant called `rename_with()`, which allows you to use a function to rename columns, like this.

```{r echo = TRUE}
dat |> rename_with(tolower) |> slice_head(n = 10)
```

Believe it or not, we've covered most of the commonly used functions in `dplyr`. The only ones left are the `bind_` family for rows and columns. Let's look at an example with `bind_rows`. Unlike the `rbind` function in base R, this one does not require all the columns to match to bind rows. The row binding will be the union of the two data frames, e.g. every column will be included, filled with `NA` if needed.

Here's an example of something not necessary to do of course, but gets the point across!

```{r echo = TRUE}
dat1 <- dat |> filter(!is.na(DOSE)) |> select(ID, TIME, DOSE)
dat2 <- dat |> filter(!is.na(OUT)) |> select(ID, TIME, OUT)
#look at them
dat1
dat2
#bind them and re-order
dat3 <- bind_rows(dat1, dat2) |> arrange(ID, TIME)
dat3
```

**Et voilá!**

## Data Visualization with Plotly

### Introduction

The official plotting package of the Tidyverse is `ggplot2`. There are many great tutorials online for `ggplot2`. This package popularized the concept of successively adding layers to plots to build the final product. Interestingly, the operator between layers in `ggplot2` is a `+` and not a pipe operator. While `ggplot2` is an amazing package that generates beautiful plots, we are instead going to take a look at `plotly`, which plays very nicely with both `ggplot2` and with Tidyverse as a whole, because the layers are connected with pipes.

Plotly got its start in Python and Javascript and has packages available in R (of course), MATLAB, Perl, Julia, and Arduino.

Take a look at this:

```{r echo = TRUE}
dat |> filter(!is.na(OUT)) |> plot_ly(x = ~TIME, y = ~OUT, name = ~ID) |> add_lines() |> add_markers() |> layout(showlegend = FALSE)
```

Mouse over the plot and you will discover one of the main features of plotly that distinguishes it from other graphing packages in R: interactivity! When you hit one of the plotted points with your mouse, you see the x and y values plus the ID number, since we asked to name each trace with ID.

Let's discuss some points about the above code.

* Plotly supports piping. We start with our `dat` object, filter to keep only non-missing observations (so the lines all join contiguously over time), and then call our plot function.
* Note that the base plot call in the `plotly` package is `plot_ly`.
* Because we set up our data with the pipeline, we don't have to supply it as the first argument to `plot_ly`.
* We define our x and y variables using `~` before the name of the appropriate columns.
* Plotly calls each layer a "trace". Arguments like `name` that are placed in the top level `plot_ly` call apply to all layers/traces below, unless you add `inherit = FALSE` to one of them or another argument that overwrites the base layer argument. In this case, we used `name` at the top level to split both the line and marker traces by subject ID.
* The first layer or trace we added was lines. Then we added markers. We can of course control appearances of each of these traces individually by supplying more arguments at the trace level. More on that shortly. Lines and markers are very commonly used objects in plotly plots.
* Lastly, we modified the overall plot layout, in this case by suppressing the legend. Layout is a component of almost every plotly plot as it controls every feature that isn't part of a trace, e.g. axes and legends.

Play around with moving the `name = ~ID` argument to `add_lines()`, then `add_marker()`, then both functions, and then removing it entirely. It might help to also change the layout to include `showlegend = TRUE` just to see what traces are being plotted. This code will get you started. We also show that the call to `plot_ly` always returns a plotly object, which we call p in this example. To plot it, simply send the value to the console. Add your additional lines of code to the snippet below.

```{r plotly-intro, exercise = TRUE}

p <- dat |> filter(!is.na(OUT)) |> 
  plot_ly(x = ~TIME, y = ~OUT) |> 
  add_lines() |> 
  add_markers(name = ~ID) |> 
  layout(showlegend = TRUE)

p
```

For your last code with no `name` arguments, try adding a `group_by` to your pipeline to group by ID.

```{r plotly-intro-2, exercise = TRUE}

```

```{r plotly-intro-2-solution}

p <- dat |> group_by(ID) |> filter(!is.na(OUT)) |> plot_ly(x = ~TIME, y = ~OUT) |> add_lines() |> add_markers() |> layout(showlegend = TRUE)

p
```


### Formatting elements

Plotly is built on the concept of graphical layers, as we mentioned. We've seen two layers added in the examples we've worked on: markers and lines. To exercise finer control over each layer, we can embed lists, often nested. Let's change the characteristics of the markers, lines, and axis labels.

```{r echo = TRUE}
p <- dat |> group_by(ID) |> filter(!is.na(OUT)) |> plot_ly(x = ~TIME, y = ~OUT) |> 
  add_lines(line = list(color = "dodgerblue", width = 2)) |> 
  add_markers(marker = list(color = "yellow", size = 16, symbol = "diamond", opacity = 0.5, line = list(color = "forestgreen", width = 3))) |> 
  layout(showlegend = FALSE, 
         xaxis = list(title = list(
           text = "<b>Time</b>",
           font = list(size = 20, color = "navy")
         )),
         yaxis = list(title = list(
           text = "<b>Concentration</b>",
           font = list(size = 20, color = "navy")
         ))
         )
p
```

Points to consider:

* The `line` argument for `add_lines` is a list with attributes. We've included `color` and `width`.
* Similarly, the `marker` argument for `add_markers` is a list. We've included these items in the list:
  - `color`for the color of the marker fill
  - `size` for marker size in points
  - `symbol` for the marker shape
  - `opacity` to control the alpha (transparency) of the marker, from 0 (invisible) to 1 (opaque)
  - `line` Not to be confused with the `line` argument in the `add_lines()` function, here `line` is an element in the `marker` list which controls the appearance of the outline around the marker. We gave it a "forestgreen" color and a width of 3.
* In the layout layer, we suppress the legend as before, but we also now control the appearance of the x- and y-axes using lists named as `xaxis` and `yaxis`. There are many customizations for each axis, but we focused on the `title` element of the axis list. The `title` element is itself a list.
  - `text` The axis label text. Note that we took advantage of plotly's tight integration with HTML to use HTML codes for bolding the text:`<b>` and `</b>` surrounding the text we want to bold.
  - `font` Another list within a list, here controling the appearance of the title text with `size` in points and `color`.
  
### A different plot type

Try to put all your new skills to use to create a box plot comparing drug concentrations between men and women. Be sure to mouse over the plot after you're done to see the info that plotly gives you.

**Hints:**

* Use the GENDER column in dat. First in your pipeline is to fill in missing GENDER values.
* Next in the pipeline, create a new variable called "sex" which is a factor based on GENDER with `labels = c("Male", "Female")`. If you need it, look at the help for factor in R by typing `?factor` into the console window of Rstudio and pushing \<Return\>.
* Next, filter out missing OUT values. We've done this previously.
* Now call the plot with `plot_ly`. Define your x and y values. Separate the trace by your newly created `sex` factor using the `name` argument.
* Append an `add_boxplot()` function to your pipeline.
* Lastly, modify the layout just as we did before, suppressing the legend and naming the axes appropriately. You can remove the bold tags.

```{r plotly-boxplot, exercise = TRUE}

```
  
```{r plotly-boxplot-solution}
dat |> 
  fill(GENDER) |> #fill in missing values
  mutate(sex = factor(GENDER, labels = c("Male", "Female"))) |> #create a new factor
  filter(!is.na(OUT)) |> #focus on non-missing observations
  plot_ly(x = ~sex, y = ~OUT, name = ~sex) |>  #set up the plot
  add_boxplot() |> #add a boxplot layer (trace)
  layout(showlegend = FALSE, #suppress the legend
         xaxis = list(title = list( #modify the axes labels similarly to before
           text = "Sex",
           font = list(size = 20, color = "navy")
         )),
         yaxis = list(title = list(
           text = "Concentration",
           font = list(size = 20, color = "navy")
         ))
         )

```
  
There are endless ways to customize plotly plots. Here are resources to help:

* `schema()` This function typed into your R console opens a tree with the current options for all plots. Levels in the tree can be accessed by expanding the `r knitr::asis_output("\U25B6")` next to each element. The main two elements to explore are below.
  - **traces** Expand to see all the layers / traces that can be added. The most commonly used is **scatter**, and the most common heading to expand under that is *attributes*. Most common attributes are line, marker, and name.
  - **layout** Expand *layoutAttributes*. Most commonly used are title, xaxis, yaxis, annotations (for adding labels), legend, shapes, and showlegend.
* Help for the following functions: 
  - `plot_ly` The help for this function shows all the top level arguments that will be applied to every trace. In many cases these top level arguments map to arguments in a trace, even though the name may be different. For example, `stroke` maps to the color of the outline for markers (`marker.line.color`) and the color of lines (`line.color`). 
  - `add_lines` This help page will show you all the `add_*()` functions available, although the actual documentation for each function is sparse and not very helpful. It's more to indicate which functions are available. Use `schema()` to understand what arguments you can supply to these functions, e.g. `r knitr::asis_output("\U25B6")` traces `r knitr::asis_output("\U25B6")` scatter `r knitr::asis_output("\U25B6")` attributes `r knitr::asis_output("\U25B6")` line to know what arguments you can use with `add_lines` or `r knitr::asis_output("\U25B6")` traces `r knitr::asis_output("\U25B6")` scatter `r knitr::asis_output("\U25B6")` attributes `r knitr::asis_output("\U25B6")` marker to know what arguments you can use with `add_markers`. Note that if you add some arguments here, they will likely override any arguments that you might have used in the top level `plot_ly` call.
* The online reference book for [plotly in R](https://plotly-r.com/index.html){target="_blank"} is a good primer.


**Lastly**, if you are familiar with `ggplot2`, you can often simply transform your ggplot into an interactive plotly plot and have the best of both worlds!

```{r echo = TRUE}
p <- dat |> filter(!is.na(OUT)) |> mutate(ID = factor(ID)) |>
  ggplot(aes(x = TIME, y = OUT, group = ID, color = ID)) + geom_point() + geom_line() + theme(legend.position = "none")
p #static ggplot
ggplotly(p, tooltip = c("x", "y", "color")) #dynamic plotly plot
```



